Component response rate variation underlies the stability of complex systems

### 2020-03-11 notebook/ms.tex
Change the notation of minimum real part of eigenvalues

Randomly assembled complex systems can be represented as large square
matrices (\(\mathbf{M}\)) with \(S\) components (e.g., networks of
species\textsuperscript{\protect\hyperlink{ref-Allesina2012}{2}} or
banks\textsuperscript{\protect\hyperlink{ref-Haldane2011}{11}}). One
element of such a matrix, \(M_{ij}\), defines how component \(j\)
affects component \(i\) in the system at a point of
equilibrium\textsuperscript{\protect\hyperlink{ref-Allesina2012}{2}}.
Off-diagonal elements (\(i \neq j\)) therefore define interactions
between components, while diagonal elements (\(i = j\)) define component
self-regulation (\(d\), e.g., carrying capacity in ecological
communities). Traditionally, off-diagonal elements are assigned non-zero
values with a probability \(C\), which are sampled from a distribution
with variance \(\sigma_{M}^{2}\); diagonal elements are set to
\(d = -1\)\textsuperscript{\protect\hyperlink{ref-May1972}{1},\protect\hyperlink{ref-Allesina2012}{2},\protect\hyperlink{ref-Allesina2015}{4}}.
Local system stability is assessed using eigenanalysis, with the system
being stable if the real parts of all eigenvalues (\(\lambda\)) of
\(\mathbf{M}\) are negative
(\(\max\left(\Re(\lambda)\right) < 0\))\textsuperscript{\protect\hyperlink{ref-May1972}{1},\protect\hyperlink{ref-Allesina2012}{2}}.
In a large system (high \(S\)), eigenvalues are distributed
uniformly\textsuperscript{\protect\hyperlink{ref-Tao2010}{14}} within a
circle centred at \(\Re = -1\) (the mean value of diagonal elements) and
\(\Im = 0\), with a radius of
\(\sigma_{M}\sqrt{SC}\)\textsuperscript{\protect\hyperlink{ref-May1972}{1},\protect\hyperlink{ref-Allesina2012}{2},\protect\hyperlink{ref-Allesina2015}{4}}
(Figs 1a and 2a). Local stability of randomly assembled systems
therefore becomes increasingly unlikely as \(S\), \(C\), and
\(\sigma_{M}\) increase.

@@

\textbf{Effect of \(\mathbf{\sigma^{2}_{\gamma}}\) on \(\mathbf{M}\)
(co)variation}. The value of \(\max(\Re(\lambda))\), and therefore
system stability, can be estimated from five properties of
\(\mathbf{M}\)\textsuperscript{\protect\hyperlink{ref-Tang2014b}{19}}.
These properties include (1) system size (\(S\)), (2) mean
self-regulation of components (\(d\)), (3) mean interaction strength
between components (\(\mu\)), (4) the standard deviation between
component interaction strengths (\(\sigma_{M}\)), and (5) the
correlation of interaction strengths between components, \(M_{ij}\) and
\(M_{ji}\) (\(\rho\)). Positive \(\sigma_{\gamma}\) does not change
\(S\), nor does it necessarily change \(E[d]\) or \(E[\mu]\). What
\(\sigma_{\gamma}\) does change is the total variation in component
interaction strengths (\(\sigma_{M}\)), and \(\rho\). Introducing
variation in \(\gamma\) increases the total variation in the system.
Variation of the off-diagonal elements in \(\mathbf{M}\) is described by
the joint variation of two random variables,

@@

Given \(\sigma^{2}_\gamma = 0\), \(\max(\Re(\lambda))\) decrease
linearly with \(\rho\) such
that\textsuperscript{\protect\hyperlink{ref-Tang2014c}{23}},

@@

Ten thousand numerical simulations of \(\mathbf{M}\) revealed that
\(\sigma^{2}_{\gamma} > 0\) results in a nonlinear relationship between
\(\rho\) and \(\max(\Re(\lambda))\), which can sometimes increase the
stability of the system. Figure 2 shows a comparison of
\(\max(\Re(\lambda))\) across \(\rho\) values when
\(\sigma^{2}_{\gamma} = 0\) versus \(\sigma^{2}_{\gamma} = 1/3\) given
\(S = 25\) and \(\sigma = 0.4\). For \(-0.4 \leq \rho \leq 0.7\) (shaded
region of Figure 2), expected \(\max(\Re(\lambda))\) was lower given
\(\sigma^{2}_{\gamma} = 1/3\). For \(\rho \geq -0.1\), the range of
\(\max(\Re(\lambda))\) values observed given
\(\sigma^{2}_{\gamma} = 1/3\) also had a lower bound, which included
negative values for \(\rho = -0.1\) and \(\rho = 0\). Hence, across a
wide range of system correlations, variation in the response rate of
system components had a stabilising effect.

The stabilising effect of \(\sigma^{2}_{\gamma}\) across \(\rho\)
increased with increasing \(S\). Figure 3 shows 10000 numerical
simulations of \(\mathbf{M}\) for \(\sigma^{2}_{\gamma} = 0\) versus
\(\sigma^{2}_{\gamma} = 1/3\) across increasing \(S\) given
\(\sigma = 0.2\) (\(\sigma\) has been lowered here to better illustrate
the effect of \(S\); note that now given \(S = 25\),
\(1 = \sigma\sqrt{SC}\)). For relatively small systems (\(S \leq 25\)),
positive \(\sigma^{2}_{\gamma}\) never decreased the expected
\(\max(\Re(\lambda))\). But as \(S\) increased, the curvilinear
relationship between \(\rho\) and \(\max(\Re(\lambda))\) resulted in
\(\mathbf{M}\) with lower expected \(\max(\Re(\lambda))\) for low
magnitudes of \(\rho\). In turn, as \(S\) increased, and systems became
more complex, positive \(\sigma^{2}_{\gamma}\) increased the proportion
of numerical simulations that were observed to be stable (see below).

@@

\textbf{Targeted manipulation of \(\mathbf{\gamma}\)}. To further
investigate the potential of \(Var(\gamma)\) to be stabilising, I used a
genetic algorithm. Genetic algorithms are heuristic tools that mimic
evolution by natural selection, and are useful when the space of
potential solutions (in this case, possible combinations of \(\gamma\)
values leading to stability in a complex system) is too large to search
exhaustively\textsuperscript{\protect\hyperlink{ref-Hamblin2013}{24}}.
Generations of selection on \(\gamma\) value combinations to minimise
\(\max\left(\Re(\lambda)\right)\) demonstrated the potential for
\(Var(\gamma)\) to increase system stability. Across
\(S = \{2, 3, ..., 39, 40\}\), sets of \(\gamma\) values were found that
resulted in stable systems with probabilities that were up to four
orders of magnitude higher than when \(\gamma = 1\) (Fig. 4), meaning
that stability could often be achieved by manipulating \(S\) \(\gamma\)
values rather than \(S \times S\) \(\mathbf{M}\) elements (i.e., by
manipulating component response rates rather than interactions between
components).

@@

For each \(S = \{2, 3, ..., 39, 40\}\), the genetic algorithm was run
for 100000 random \(\mathbf{M}\) (\(\sigma = 0.4\), \(C = 1\)). The
genetic algorithm was initialised with a population of 1000 different
\(\mathbf{\gamma}\) vectors with elements sampled i.i.d from
\(\gamma_{i} \sim \mathcal{U}(0, 2)\). Eigenanalysis was performed on
the \(\mathbf{M}\) resulting from each \(\mathbf{\gamma}\) vector, and
the 20 \(\mathbf{\gamma}\) vectors resulting in \(\mathbf{M}\) with the
lowest \(\max\left(\Re(\lambda)\right)\) each produced 50 clonal
offspring with subsequent random mutation and crossover between the
resulting new generation of 1000 \(\mathbf{\gamma}\) vectors. Mutation
of each \(\gamma_{i}\) in a \(\mathbf{\gamma}\) vector occurred with a
probability of 0.2, resulting in a mutation effect of size
\(\mathcal{N}(0, 0.02)\) being added to generate the newly mutated
\(\gamma_{i}\) (any \(\gamma_{i}\) values that mutated below zero were
multiplied by \(-1\), and any values that mutated above 2 were set to
2). Crossover occurred between two sets of 100 \(\mathbf{\gamma}\)
vectors paired in each generation; vectors were randomly sampled with
replacement among but not within sets. Vector pairs selected for
crossover swapped all elements between and including two \(\gamma_{i}\)
randomly selected with replacement (this allowed for reversal of vector
element positions during crossover; e.g.,
\(\{\gamma_{4}, \gamma_{5}, \gamma_{6}, \gamma_{7}\} \to \{\gamma_{7}, \gamma_{6}, \gamma_{5}, \gamma_{4}\}\)
). The genetic algorithm terminated if a stable \(\mathbf{M}\) was
found, 20 generations occurred, or if the mean \(\mathbf{\gamma}\)
fitness increase between generations was less than 0.01 (where fitness
was defined as \(W_{\gamma} = -\max\left(\Re(\lambda)\right)\) for
\(\mathbf{M}\)).

@@

\textbf{Figure 1: Eigenvalue distributions of random complex systems.}
Each panel shows the real (x-axis) and imaginary (y-axis) parts of
\(S =\) 400 eigenvalues from random \(S \times S\) matrices.
(\(\textbf{a}\)) A system represented by a matrix \(\mathbf{A}\), in
which all elements are sampled from a normal distribution with
\(\mu = 0\) and \(\sigma_{A} = 1/\sqrt{S}\). Points are uniformly
distributed within a circle centred at the origin with a radius of
\(\sigma_{A} \sqrt{S} = 1\). (\(\textbf{b}\)) The same system after
including variation in the response rates of \(S\) components,
represented by the diagonal matrix \(\gamma\), such that
\(\mathbf{M} = \gamma\mathbf{A}\). Elements of \(\gamma\) are randomly
sampled from a uniform distribution from \(\min = 0\) to \(\max = 2\).
Eigenvalues of \(\mathbf{M}\) are then distributed non-uniformly within
a circle centred at the origin with a radius of
\(\sqrt{\sigma^{2}_{A}(1 + \sigma^{2}_{\gamma})S} \approx\) 1.17.
(\(\textbf{c}\)) A different random system \(\mathbf{A}\) constructed
from the same parameters as in \(\textbf{a}\), except with diagonal
values of \(-1\). (\(\textbf{d}\)) The same system \(\textbf{c}\) after
including variation in component response rates, sampled from
\(\mathcal{U}(0, 2)\) as in \(\textbf{b}\).

>>>>

Randomly assembled complex systems can be represented as large square
matrices (\(\mathbf{M}\)) with \(S\) components (e.g., networks of
species\textsuperscript{\protect\hyperlink{ref-Allesina2012}{2}} or
banks\textsuperscript{\protect\hyperlink{ref-Haldane2011}{11}}). One
element of such a matrix, \(M_{ij}\), defines how component \(j\)
affects component \(i\) in the system at a point of
equilibrium\textsuperscript{\protect\hyperlink{ref-Allesina2012}{2}}.
Off-diagonal elements (\(i \neq j\)) therefore define interactions
between components, while diagonal elements (\(i = j\)) define component
self-regulation (\(d\), e.g., carrying capacity in ecological
communities). Traditionally, off-diagonal elements are assigned non-zero
values with a probability \(C\), which are sampled from a distribution
with variance \(\sigma_{M}^{2}\); diagonal elements are set to
\(d = -1\)\textsuperscript{\protect\hyperlink{ref-May1972}{1},\protect\hyperlink{ref-Allesina2012}{2},\protect\hyperlink{ref-Allesina2015}{4}}.
Local system stability is assessed using eigenanalysis, with the system
being stable if the real parts of all eigenvalues (\(\lambda\)) of
\(\mathbf{M}\) are negative
(\(\Re(\lambda_{max}) < 0\))\textsuperscript{\protect\hyperlink{ref-May1972}{1},\protect\hyperlink{ref-Allesina2012}{2}}.
In a large system (high \(S\)), eigenvalues are distributed
uniformly\textsuperscript{\protect\hyperlink{ref-Tao2010}{14}} within a
circle centred at \(\Re = -1\) (the mean value of diagonal elements) and
\(\Im = 0\), with a radius of
\(\sigma_{M}\sqrt{SC}\)\textsuperscript{\protect\hyperlink{ref-May1972}{1},\protect\hyperlink{ref-Allesina2012}{2},\protect\hyperlink{ref-Allesina2015}{4}}
(Figs 1a and 2a). Local stability of randomly assembled systems
therefore becomes increasingly unlikely as \(S\), \(C\), and
\(\sigma_{M}\) increase.

@@

\textbf{Effect of \(\mathbf{\sigma^{2}_{\gamma}}\) on \(\mathbf{M}\)
(co)variation}. The value of \(\Re(\lambda_{max})\), and therefore
system stability, can be estimated from five properties of
\(\mathbf{M}\)\textsuperscript{\protect\hyperlink{ref-Tang2014b}{19}}.
These properties include (1) system size (\(S\)), (2) mean
self-regulation of components (\(d\)), (3) mean interaction strength
between components (\(\mu\)), (4) the standard deviation between
component interaction strengths (\(\sigma_{M}\)), and (5) the
correlation of interaction strengths between components, \(M_{ij}\) and
\(M_{ji}\) (\(\rho\)). Positive \(\sigma_{\gamma}\) does not change
\(S\), nor does it necessarily change \(E[d]\) or \(E[\mu]\). What
\(\sigma_{\gamma}\) does change is the total variation in component
interaction strengths (\(\sigma_{M}\)), and \(\rho\). Introducing
variation in \(\gamma\) increases the total variation in the system.
Variation of the off-diagonal elements in \(\mathbf{M}\) is described by
the joint variation of two random variables,

@@

Given \(\sigma^{2}_\gamma = 0\), \(\Re(\lambda_{max})\) decrease
linearly with \(\rho\) such
that\textsuperscript{\protect\hyperlink{ref-Tang2014c}{23}},

@@

Ten thousand numerical simulations of \(\mathbf{M}\) revealed that
\(\sigma^{2}_{\gamma} > 0\) results in a nonlinear relationship between
\(\rho\) and \(\Re(\lambda_{max})\), which can sometimes increase the
stability of the system. Figure 2 shows a comparison of
\(\Re(\lambda_{max})\) across \(\rho\) values when
\(\sigma^{2}_{\gamma} = 0\) versus \(\sigma^{2}_{\gamma} = 1/3\) given
\(S = 25\) and \(\sigma = 0.4\). For \(-0.4 \leq \rho \leq 0.7\) (shaded
region of Figure 2), expected \(\Re(\lambda_{max})\) was lower given
\(\sigma^{2}_{\gamma} = 1/3\). For \(\rho \geq -0.1\), the range of
\(\Re(\lambda_{max})\) values observed given
\(\sigma^{2}_{\gamma} = 1/3\) also had a lower bound, which included
negative values for \(\rho = -0.1\) and \(\rho = 0\). Hence, across a
wide range of system correlations, variation in the response rate of
system components had a stabilising effect.

The stabilising effect of \(\sigma^{2}_{\gamma}\) across \(\rho\)
increased with increasing \(S\). Figure 3 shows 10000 numerical
simulations of \(\mathbf{M}\) for \(\sigma^{2}_{\gamma} = 0\) versus
\(\sigma^{2}_{\gamma} = 1/3\) across increasing \(S\) given
\(\sigma = 0.2\) (\(\sigma\) has been lowered here to better illustrate
the effect of \(S\); note that now given \(S = 25\),
\(1 = \sigma\sqrt{SC}\)). For relatively small systems (\(S \leq 25\)),
positive \(\sigma^{2}_{\gamma}\) never decreased the expected
\(\Re(\lambda_{max})\). But as \(S\) increased, the curvilinear
relationship between \(\rho\) and \(\Re(\lambda_{max})\) resulted in
\(\mathbf{M}\) with lower expected \(\Re(\lambda_{max})\) for low
magnitudes of \(\rho\). In turn, as \(S\) increased, and systems became
more complex, positive \(\sigma^{2}_{\gamma}\) increased the proportion
of numerical simulations that were observed to be stable (see below).

@@

\textbf{Targeted manipulation of \(\mathbf{\gamma}\)}. To further
investigate the potential of \(Var(\gamma)\) to be stabilising, I used a
genetic algorithm. Genetic algorithms are heuristic tools that mimic
evolution by natural selection, and are useful when the space of
potential solutions (in this case, possible combinations of \(\gamma\)
values leading to stability in a complex system) is too large to search
exhaustively\textsuperscript{\protect\hyperlink{ref-Hamblin2013}{24}}.
Generations of selection on \(\gamma\) value combinations to minimise
\(\Re(\lambda_{max})\) demonstrated the potential for \(Var(\gamma)\) to
increase system stability. Across \(S = \{2, 3, ..., 39, 40\}\), sets of
\(\gamma\) values were found that resulted in stable systems with
probabilities that were up to four orders of magnitude higher than when
\(\gamma = 1\) (Fig. 4), meaning that stability could often be achieved
by manipulating \(S\) \(\gamma\) values rather than \(S \times S\)
\(\mathbf{M}\) elements (i.e., by manipulating component response rates
rather than interactions between components).

@@

For each \(S = \{2, 3, ..., 39, 40\}\), the genetic algorithm was run
for 100000 random \(\mathbf{M}\) (\(\sigma = 0.4\), \(C = 1\)). The
genetic algorithm was initialised with a population of 1000 different
\(\mathbf{\gamma}\) vectors with elements sampled i.i.d from
\(\gamma_{i} \sim \mathcal{U}(0, 2)\). Eigenanalysis was performed on
the \(\mathbf{M}\) resulting from each \(\mathbf{\gamma}\) vector, and
the 20 \(\mathbf{\gamma}\) vectors resulting in \(\mathbf{M}\) with the
lowest \(\Re(\lambda_{max})\) each produced 50 clonal offspring with
subsequent random mutation and crossover between the resulting new
generation of 1000 \(\mathbf{\gamma}\) vectors. Mutation of each
\(\gamma_{i}\) in a \(\mathbf{\gamma}\) vector occurred with a
probability of 0.2, resulting in a mutation effect of size
\(\mathcal{N}(0, 0.02)\) being added to generate the newly mutated
\(\gamma_{i}\) (any \(\gamma_{i}\) values that mutated below zero were
multiplied by \(-1\), and any values that mutated above 2 were set to
2). Crossover occurred between two sets of 100 \(\mathbf{\gamma}\)
vectors paired in each generation; vectors were randomly sampled with
replacement among but not within sets. Vector pairs selected for
crossover swapped all elements between and including two \(\gamma_{i}\)
randomly selected with replacement (this allowed for reversal of vector
element positions during crossover; e.g.,
\(\{\gamma_{4}, \gamma_{5}, \gamma_{6}, \gamma_{7}\} \to \{\gamma_{7}, \gamma_{6}, \gamma_{5}, \gamma_{4}\}\)
). The genetic algorithm terminated if a stable \(\mathbf{M}\) was
found, 20 generations occurred, or if the mean \(\mathbf{\gamma}\)
fitness increase between generations was less than 0.01 (where fitness
was defined as \(W_{\gamma} = -\Re(\lambda_{max})\) for \(\mathbf{M}\)).

@@

\textbf{Figure 1: Eigenvalue distributions of random complex systems.}
Each panel shows the real (x-axis) and imaginary (y-axis) parts of
\(S =\) 400 eigenvalues from random \(S \times S\) matrices.
(\(\textbf{a}\)) A system represented by a matrix \(\mathbf{A}\), in
which all elements are sampled from a normal distribution with
\(\mu = 0\) and \(\sigma_{A} = 1/\sqrt{S}\). Points are uniformly
distributed within a circle centred at the origin with a radius of
\(\sigma_{A} \sqrt{S} = 1\). (\(\textbf{b}\)) The same system after
including variation in the response rates of \(S\) components,
represented by the diagonal matrix \(\gamma\), such that
\(\mathbf{M} = \gamma\mathbf{A}\). Elements of \(\gamma\) are randomly
sampled from a uniform distribution from \(\min = 0\) to \(\max = 2\).
Eigenvalues of \(\mathbf{M}\) are then distributed non-uniformly within
a circle centred at the origin with a radius of
\(\sqrt{\sigma^{2}_{A}(1 + \sigma^{2}_{\gamma})S} \approx\) 1.18.
(\(\textbf{c}\)) A different random system \(\mathbf{A}\) constructed
from the same parameters as in \(\textbf{a}\), except with diagonal
values of \(-1\). (\(\textbf{d}\)) The same system \(\textbf{c}\) after
including variation in component response rates, sampled from
\(\mathcal{U}(0, 2)\) as in \(\textbf{b}\).
